Reproducible development environments with rix

### 2024-01-30 paper.qmd
typo
review and cite dolstra

If an author provides this information, other people trying to reproduce the
study (or the author him- or herself in the future) can read this file and see
which version of [R]{.proglang} was used, and which packages (and their
versions) were used as well. However, others would still need to install the
correct software themselves. An alternative to this is instead to use the
[renv]{.pkg} package which generates a so-called `renv.lock` file which also
lists [R]{.proglang} and package versions. Here is an example of such an
`renv.lock` file:

@@

This file lists every package alongside their versions and the repository from
which they were downloaded. Generating this file only requires one to run the
`renv::init()` function. Someone else can then restore the same package library
by running `renv::restore()`. The exact same packages get installed in an
isolated, project-specific, library which doesn't interfere with the main
library of the user. [renv]{.pkg} does not restore the [R]{.proglang} version
used though, so installing the right version of R needs to be handled
separately. Before continuing, it should be noted that other packages exist
which provide similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg}
by @simonsohn2023 which makes it rather easy to install packages as they were on
CRAN at a given date. For example, the code snippet below installs the
[purrr]{.pkg} and [ggplot2]{.pkg} packages as they were on April 4th, 2017:

@@

```
options(repos = c(REPO_NAME = "https://packagemanager.posit.co/cran/
                               __linux__/jammy/2023-06-30"))
```

@@

The next step in reaching the gold standard of reproducibility would be to not
only install the right packages used for the analysis, but also the right
version of [R]{.proglang}. Of course, it would be possible to install the right
version manually, but here too, there are tools that simplify the process. One
such tool is [rig]{.proglang} by the R infrastructure team [-@rlib2023].
[rig]{.proglang} makes it easy to install different versions of [R]{.proglang},
so one could use it to install the right version to reproduce an analysis, and
then use one of the listed packages above to install the right library of of
[R]{.proglang} packages. However, this involves many manual steps and is thus
error prone. It is also time-consuming to do so, especially if one wants to
reproduce many studies.

The final tool one could use to reach the gold standard of reproducibility is to
package the right version of [R]{.proglang} and [R]{.proglang} packages inside a
[Docker]{.proglang} image. [Docker]{.proglang} is a containerisation tool: using
[Docker]{.proglang} it is possible to package some *data product* with its
dependencies into an image. A statistical analysis, from the simplest to the
most complex one, can be seen as such a data product that comes with many
software dependencies. *Dockerizing* an analysis consists in first building an
image: in the build step, the dependencies of the analysis have to be installed
and this can be achieved using the tools mentioned above. The scripts to run the
analysis and the data are also added to the image at build time. Instances of
that image, called containers, can then be executed, which give access to the
exact development environment originally used for the study. This execution step
is called the "run step". During this run step, the analysis scripts can be
executed to product the output.

The advantage of using [Docker]{.proglang} is not only about having the right
version of [R]{.proglang} and [R]{.proglang} packages bundled alongside the
analysis scripts: [Docker]{.proglang} images are actually almost complete
operating systems, which means that system-level dependecies also get bundled
into the image. This is especially useful if packages that rely on external
libraries get used for the analysis, as these external libraries are thus also
made available to future reproducers of the study.

@@

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then [Docker]{.proglang} for all the other
software, including the programming language that was used. Ideally, a build
automation tool such as [make]{.proglang} should be added to the list. It should
also be noted that these requirements are not limited to the [R]{.proglang}
programming language. A very similar approach should be taken if one uses
[Python]{.proglang} for statistical analysis instead. Furthermore, the previous
discussion also implies that for an analysis to actually be reproducible at all,
it must be made using free and open source software, as proprietary software
cannot be freely shared with others that wish to reproduce a study, and there's
no guarantee that others will have access to the proprietary software used.

As an alternative to the above approach, we will present the [Nix]{.proglang}
package manager, which is available for all major operating systems and focuses
on installing and building software in a reproducible manner. This means that
this single tool can replace not only [Docker]{.proglang} but also [renv]{.pkg}
and a build automation tool such as [Make]{.proglang}. To make [Nix]{.proglang}
more accessible to [R]{.proglang} programmers, we developed the [rix]{.pkg}
package which we will present in this article.

@@

But why is installing software with [Nix]{.proglang} reproducible? Because the
whole set of [Nix]{.proglang} expressions is hosted on Github, it is possible to
use a specific commit of `nixpkgs` to ensure reproducibility of our project.
Using a specific commit of `nixpkgs` is called *pinning a revision (or commit)*.
Pinning a revision ensures that every package that Nix installs will always be
at exactly the same versions, regardless of when in the future the packages get
built. This is because the [Nix]{.proglang} expressions that get downloaded when
pinning are the expressions as they were at that commit of `nixpkgs`.

@@

```{r, include = FALSE}
set.seed(1234)
out_r <- sample(seq(1, 10), 5)
set.seedt(NULL)
```

@@

>>>>

If an author provides this information, other people trying to reproduce the
study (or the author him- or herself in the future) can read this file and see
which version of [R]{.proglang} was used, and which packages (and their
versions) were used as well. However, others would still need to install the
correct versions of each package themselves. An alternative to this is instead
to use the [renv]{.pkg} package which generates a so-called `renv.lock` file
which also lists [R]{.proglang} and package versions. Here is an example of such
an `renv.lock` file:

@@

This file lists every package alongside their versions and the repository from
which they were downloaded. Generating this file only requires one to run the
`renv::init()` function. Someone else can then restore the same package library
by running `renv::restore()`. The exact same packages get installed in an
isolated, project-specific library which doesn't interfere with the main
library of the user. [renv]{.pkg} does not restore the [R]{.proglang} version
used though, so installing the right version of R needs to be handled
separately. Before continuing, it should be noted that other packages exist
which provide similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg}
by @simonsohn2023 which makes it rather easy to install packages as they were on
CRAN at a given date. For example, the code snippet below installs the
[purrr]{.pkg} and [ggplot2]{.pkg} packages as they were on April 4th, 2017:

@@

```
options(repos =
  c(REPO_NAME =
    "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
  )
)
```

@@

The next step in reaching the gold standard of reproducibility would be to not
only install the right packages used for the analysis, but also the right
version of [R]{.proglang}. Of course, it would be possible to install the right
version manually, but here too, there are tools that simplify the process. One
such tool is [rig]{.proglang} by the R infrastructure team [-@rlib2023].
[rig]{.proglang} makes it easy to install different versions of [R]{.proglang},
so one could use it to install the right version to reproduce an analysis, and
then use one of the listed packages above to install the right library of
[R]{.proglang} packages. However, this involves many manual steps and is thus
error prone. It is also time-consuming to do so, especially if one wants to
reproduce many studies.

The final tool one could use to reach the gold standard of reproducibility is to
package the right version of [R]{.proglang} and [R]{.proglang} packages inside a
[Docker]{.proglang} image. [Docker]{.proglang} is a containerisation tool: using
[Docker]{.proglang} it is possible to package some *data product* with its
dependencies into an image. A statistical analysis, from the simplest to the
most complex one, can be seen as such a data product that requires many software
dependencies. *Dockerizing* an analysis consists in first building an image: in
the build step, the dependencies of the analysis have to be installed and this
can be achieved using the tools mentioned above. The scripts to run the analysis
and the data are also added to the image at build time. It is possible to then
execute the code of a study inside a so-called container, which is an instance
of an image. Executing, or running, these containers is done non-interactively,
and so it is possible to instantiate the correct environment to run a study in a
single command.

The advantage of using [Docker]{.proglang} is not only about having the right
version of [R]{.proglang} and [R]{.proglang} packages bundled alongside the
analysis scripts: [Docker]{.proglang} images are actually almost complete
operating systems, which means that system-level dependecies also get bundled
into the image. This is especially useful for packages that rely on external
libraries to run, as these external libraries are thus also made available to
future reproducers of the study within the [Docker]{.proglang} image. The image
can also be shared.

@@

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then [Docker]{.proglang} for all the other
software, including the programming language that was used. Ideally, a build
automation tool such as [Make]{.proglang} should be added to the list. It should
also be noted that these requirements are not limited to the [R]{.proglang}
programming language. A very similar approach should be taken if one uses
[Python]{.proglang} for statistical analysis instead. Furthermore, the previous
discussion also implies that for an analysis to actually be reproducible at all,
it must be made using free and open source software, as proprietary software
cannot be freely shared with others that wish to reproduce a study, and there's
no guarantee that others will have access to the proprietary software used.

As an alternative to the above approach, we will present the [Nix]{.proglang}
package manager, which is available for all major operating systems and focuses
on installing and building software in a reproducible manner. This means that
this single tool can replace not only [Docker]{.proglang} but also [renv]{.pkg}
and even a build automation tool such as [Make]{.proglang}. To make
[Nix]{.proglang} more accessible to [R]{.proglang} programmers, we developed the
[rix]{.pkg} package which we will present in this article.

@@

But why is installing software with [Nix]{.proglang} reproducible? Because the
whole set of [Nix]{.proglang} expressions is hosted on Github, it is possible to
use a specific commit of `nixpkgs` to ensure reproducibility of our project.
Using a specific commit of `nixpkgs` is called *pinning a revision (or commit)*.
Pinning a revision ensures that every package that Nix installs will always be
at exactly the same versions, regardless of when in the future the packages get
built. This is because the [Nix]{.proglang} expressions that get downloaded when
pinning are the expressions as they were at that commit of `nixpkgs`. This is
not the only reason [Nix]{.proglang} is very useful for reproducibility, but the
[Nix]{.proglang} package manager itself was designed with reproducibility in
mind. [Nix]{.proglang} is functional package manager, meaning that it uses
concepts from functional programming to build software: no-side effects or
global variables are allowed, and the output is always the same if given the
same inputs, regardless of where or when a package gets built.
We refer to @dolstra2004nix for a more technical discussion on [Nix]{.proglang}.

@@

```{r, include = FALSE}
set.seed(1234)
out_r <- sample(seq(1, 10), 5)
set.seed(NULL)
```

@@

It is possible to use `with_nix()` from an R session that was installed itself
via Nix, or from a regular R installation.
